Simplified and Verified: A Fresh Look at a Proof-Producing Union-Find Algorithm

### 2024-08-23 paper/CPP/paper.tex
Finish section on Explain_Simple

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,typedef,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and,fixes},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
\newcommand{\lstinlinew}[2][]{\lstinline[#1]{#2}}
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

Building on this formalisation, we encapsulate this union-find data structure into an abstract data type (ADT) \lstinline|ufa| by defining
\begin{lstlisting}
typedef ufa = {uf. ufa_invar uf}
\end{lstlisting}
By lifting operations on the underlying list to the ADT using Isabelle's lifting infrastructure~\cite{lifting_transfer}, we obtain the operations
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ :: ufa $\Rightarrow$ nat rel!,
  \item \lstinline!ufa_rep_of :: ufa $\Rightarrow$ nat $\Rightarrow$ nat!, and
  \item \lstinline!ufa_union :: ufa $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ ufa!,
\end{itemize}
where \lstinline!nat rel = (nat $\times$ nat) set!.
The intended meaning of these operations is the following:
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ uf! is the partial equivalence relation represented by \lstinline!uf!,
  \item \lstinline!ufa_rep_of uf x! is the representative of the equivalence class that \lstinline!x! belongs to, and
  \item \lstinline!ufa_union uf x y! returns a union-find data structure where the equivalence classes of \lstinline!x! and \lstinline!y! are merged.
    This is implemented by updating the underlying list at index \lstinline!rep_of l x! with \lstinline|rep_of l y|.
\end{itemize}
More rigorously, the above operations fulfil the following properties:
\begin{lstlisting}
lemma ufa_rep_of uf x = ufa_rep_of uf y $\longleftrightarrow$
  (x, y) $\in$ ufa_$\alpha$ uf

lemma ufa_$\alpha$ (ufa_union uf x y) =
  per_union (ufa_$\alpha$ uf) x y
\end{lstlisting}
where \lstinline!per_union R x y! is defined as the equivalence relation that results from merging the respective equivalence classes in the relation \lstinline|R| that \lstinline!x! and \lstinline!y! belong to.

\section{Simple Certifying Union-Find Algorithm}
\begin{figure}
  \begin{gather*}
    \inferrule*[right=Assm]{\hbox{\lstinline|i < length us|} \\ \hbox{\lstinline|us ! i = (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ AssmP i : (x, y)|}} \\[0.5em]
    \inferrule*[right=Refl]{ }{\hbox{\lstinline|us $\ \vdash_=\ $ ReflP x : (x, x)|}} \\[0.5em]
    \inferrule*[right=Sym]{\hbox{\lstinline|us $\ \vdash_=\ $ p : (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ SymP p : (y, x)|}} \\[0.5em]
    \inferrule*[right=Trans]{\hbox{\lstinline|us $\ \vdash_=\ $ p1 : (x, y)|} \\ \hbox{\lstinline|us $\ \vdash_=\ $ p2 (y, z)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ p1 $\ \TransP\ $ p2|}}
  \end{gather*}
  \caption{The data type \lstinline|eq_prf| of certificates and its corresponding system of inference rules $\vdash_=$.\label{fig:eq_prf}}
\end{figure}
Building on the union-find ADT from the previous section, we now develop a simple \lstinline|explain| operation that,
for a given list of equations \lstinline|us|, takes two elements \lstinline|x| and \lstinline|y| and produces a certificate that \lstinline|x = y| modulo \lstinline|us|.
The certificate is given in terms of a data type \lstinline|eq_prf| with its corresponding system $\vdash_=$ of inference rules as seen in \autoref{fig:eq_prf}.
As expected, we have inference rules that utilise the reflexivity, symmetry, and transitivity of equality as well as an assumption rule.
To improve readability, we use the infix operator $\bigtriangledown$ to denote the proof term for transitivity.

\begin{figure*}
  \centering
  \begin{lstlisting}
function explain :: (nat $\times$ nat) list $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ nat eq_prf where
  explain [] x _ = ReflP x
| explain (us @ [(a, b)]) x y =
    let
      uf = ufa_unions uf_init us
      a_b_P = AssmP (length us)
    in
      if ufa_rep_of uf x = ufa_rep_of uf y then explain us x y
      else if ufa_rep_of uf x = ufa_rep_of uf a then
        explain us x a $\TransP$ a_b_P $\TransP$ explain us y b
      else
        explain us x b $\TransP$ SymP a_b_P $\TransP$ explain us a y
  \end{lstlisting}
  \caption{A simple implementation of the \lstinline|explain| operation.\label{fig:explain}}
\end{figure*}
Although \lstinline|explain| takes a list of equations \lstinline|us|, its implementation actually works with the union-find ADT \lstinline|uf| that results from applying \lstinline|us| as unions to an initial union-find data structure \lstinline|uf_init|.
Accordingly, we define
\begin{lstlisting}
definition ufa_unions =
  foldl ($\lambda$uf (x, y). ufa_union uf x y)
\end{lstlisting}
where we have \lstinline|uf = ufa_unions uf_init us| in terms of the constants above.
At last, we are ready to implement the \lstinline|explain| operation as depicted in \autoref{fig:explain}.\todo{explain with Cons instead of snoc}
The algorithm works with the assumption that the given elements \lstinline|x| and \lstinline|y| are equal modulo \lstinline|us|.
Furthermore, we assume that each element of the initial data structure \lstinline|uf_init| is equal to itself, i.e.\ it holds that \lstinline|ufa_$\alpha$ uf_init $\subseteq$ Id|.

@@

Otherwise, the list of unions must be equal to \lstinline|us @ [(a, b)]| for some \lstinline|us|, \lstinline|a|, and \lstinline|b|.
We distinguish between two cases:
\begin{itemize}
  \item The elements \lstinline|x| and \lstinline|y| are already equal modulo \lstinline|us|, so we proceed recursively with \lstinline|us|.
  \item In the case that the equation \lstinline|a = b| is necessary for \lstinline|x = y| to hold, we either have \lstinline|x = a| and \lstinline|b = y| or \lstinline|x = b| and \lstinline|a = y| modulo \lstinline|us|.
    Assuming that the former holds --- the other case is analogous --- we recursively construct the certificates for \lstinline|x = a| and \lstinline|b = y|.
    Together with the fact that \lstinline|a = b|, which we prove by assumption, we obtain \lstinline|x = y| by transitivity.
\end{itemize}

\begin{itemize}
  \item Soundness of the proof system
  \item Completeness of the algorithm
\end{itemize}

>>>>

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,typedef,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and,fixes,defines},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
\newcommand{\lstinlinew}[2][]{\lstinline[#1]{#2}}
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

Building on this formalisation, we encapsulate this union-find data structure into an abstract data type (ADT) \lstinline|ufa| by defining
\begin{lstlisting}
typedef ufa = {uf. ufa_invar uf}
\end{lstlisting}
By lifting operations on the underlying list to the ADT using Isabelle's lifting infrastructure~\cite{lifting_transfer}, we obtain the operations
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ :: ufa $\Rightarrow$ nat rel!,
  \item \lstinline!ufa_rep_of :: ufa $\Rightarrow$ nat $\Rightarrow$ nat!, and
  \item \lstinline!ufa_union :: ufa $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ ufa!,
\end{itemize}
where \lstinline!nat rel = (nat $\times$ nat) set!.
The intended meaning of these operations is the following:
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ uf! is the partial equivalence relation represented by \lstinline!uf!,
  \item \lstinline!ufa_rep_of uf x! is the representative of the equivalence class that \lstinline!x! belongs to, and
  \item \lstinline!ufa_union uf x y! returns a union-find data structure where the equivalence classes of \lstinline!x! and \lstinline!y! are merged.
    This is implemented by updating the underlying list at index \lstinline!rep_of l x! with \lstinline|rep_of l y|.
\end{itemize}
More rigorously, the above operations fulfil the following properties:\todo{Spacing between lemmas,defs}
\begin{lstlisting}
lemma ufa_rep_of uf x = ufa_rep_of uf y $\longleftrightarrow$
  (x, y) $\in$ ufa_$\alpha$ uf

lemma ufa_$\alpha$ (ufa_union uf x y) =
  per_union (ufa_$\alpha$ uf) x y
\end{lstlisting}
where \lstinline!per_union R x y!\todo{Definition with \lstinline|trancl|?} is defined as the equivalence relation that results from merging the respective equivalence classes in the relation \lstinline|R| that \lstinline!x! and \lstinline!y! belong to.

But what happens if \lstinline!x! or \lstinline!y! is not an element of the partial equivalence relation \lstinline|R|?\todo{Define element of per.}
In that case, the equivalence relation is unchanged, which means that \lstinline|per_union R x y = R|.
This, however, can be seen as a misuse of the union-find data structure, since we initialise it with a fixed set of elements \lstinline|A| and expect the user to only work with these elements.
Therefore, we introduce the following definitions that characterise valid union(s) with regard to the initial set \lstinline|A|.
\begin{lstlisting}
definition valid_union uf a b =
  a $\in$ Field (ufa_$\alpha$ uf) $\land$ b $\in$ Field (ufa_$\alpha$ uf)
definition valid_unions uf us =
  $\forall$(x, y) $\in$ set us. valid_union uf x y
\end{lstlisting}

\section{Simple Certifying Union-Find Algorithm}
\begin{figure}
  \begin{gather*}
    \inferrule*[right=Assm]{\hbox{\lstinline|i < length us|} \\ \hbox{\lstinline|us ! i = (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ AssmP i : (x, y)|}} \\[0.5em]
    \inferrule*[right=Refl]{ }{\hbox{\lstinline|us $\ \vdash_=\ $ ReflP x : (x, x)|}} \\[0.5em]
    \inferrule*[right=Sym]{\hbox{\lstinline|us $\ \vdash_=\ $ p : (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ SymP p : (y, x)|}} \\[0.5em]
    \inferrule*[right=Trans]{\hbox{\lstinline|us $\ \vdash_=\ $ p1 : (x, y)|} \\ \hbox{\lstinline|us $\ \vdash_=\ $ p2 (y, z)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ p1 $\ \TransP\ $ p2 : (x, z)|}}
  \end{gather*}
  \caption{The data type \lstinline|eq_prf| of certificates and its corresponding system of inference rules $\vdash_=$.\label{fig:eq_prf}}
\end{figure}
Building on the union-find ADT from the previous section, we now develop a simple \lstinline|explain| operation that,
for a given list of equations \lstinline|us|, takes two elements \lstinline|x| and \lstinline|y| and produces a certificate that \lstinline|x = y| modulo \lstinline|us|.
The certificate is given in terms of a data type \lstinline|eq_prf| with its corresponding system $\vdash_=$ of inference rules as seen in \autoref{fig:eq_prf}.
As expected, we have inference rules that utilise the reflexivity, symmetry, and transitivity of equality as well as an assumption rule.
To improve readability, we use the infix operator $\bigtriangledown$ to denote the proof term for transitivity.

We remark that one can not prove disequality with this proof system.
In that sense, we are working with an open-world assumption.

We prove that $\vdash_=$ is sound and complete with respect to the equivalence relation induced by \lstinline|us|, i.e.\ the equivalence closure of \lstinline|us|.
In Isabelle, we define\todo{Introduce inverse and rtrancl?}
\begin{lstlisting}
definition symcl r = r $\cup$ r$^{-1}$
definition equivcl r = (symcl r)$^*$
\end{lstlisting}
and prove
\begin{lstlisting}
lemma $\vdash_=$_sound:
assumes as $\vdash_=$ p : (x, y)
shows (x, y) $\in$ equivcl (set as)

lemma $\vdash_=$_complete:
assumes (x, y) $\in$ equivcl (set as)
shows $\exists$p. as $\vdash_=$ p : (x, y)
\end{lstlisting}

\begin{figure*}
  \centering
  \begin{lstlisting}
function explain :: (nat $\times$ nat) list $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ nat eq_prf where
  explain [] x _ = ReflP x
| explain (us @ [(a, b)]) x y =
    let uf = ufa_unions uf_init us; a_b_P = AssmP (length us)
    in if ufa_rep_of uf x = ufa_rep_of uf y then explain us x y
      else if ufa_rep_of uf x = ufa_rep_of uf a then explain us x a $\TransP$ a_b_P $\TransP$ explain us y b
      else explain us x b $\TransP$ SymP a_b_P $\TransP$ explain us a y

definition explain_partial us x y =
  if (x, y) $\in$ equivcl (set us) then Some (explain us x y) else None
  \end{lstlisting}
  \caption{A simple implementation of the \lstinline|explain| operation. It operates on an arbitrary initial union-find data structure \lstinline|uf_init|.\label{fig:explain}}
\end{figure*}
Although \lstinline|explain| takes a list of equations \lstinline|us|, its implementation actually works with an instance \lstinline|uf| of the union-find ADT that results from applying \lstinline|us| as unions to an initial union-find data structure \lstinline|uf_init|.
Accordingly, we define
\begin{lstlisting}
definition ufa_unions =
  foldl ($\lambda$uf (x, y). ufa_union uf x y)
\end{lstlisting}
so we have \lstinline|uf = ufa_unions uf_init us| in terms of the constants above.

At last, we are ready to implement the \lstinline|explain| operation as depicted in \autoref{fig:explain}.\todo{explain with Cons instead of snoc}
The algorithm works with the assumption that the given elements \lstinline|x| and \lstinline|y| are equal modulo \lstinline|us|.
Furthermore, we assume that each element of the initial data structure \lstinline|uf_init| is equal to itself
and that \lstinline|us| are valid unions with respect to \lstinline|uf_init|.
In formulae, we assume
\begin{itemize}
  \item \lstinline|ufa_$\alpha$ uf_init $\subseteq$ Id|, and
  \item \lstinline|valid_unions uf_init us|
\end{itemize}
in what follows.

@@

Otherwise, the list of unions must be equal to \lstinline|us @ [(a, b)]| for some \lstinline|us|, \lstinline|a|, and \lstinline|b|.
We distinguish between two cases:
\begin{itemize}
  \item The elements \lstinline|x| and \lstinline|y| are already equal modulo \lstinline|us|, so we proceed recursively with \lstinline|us|.
  \item In the case that the equation \lstinline|a = b| is necessary for \lstinline|x = y| to hold, we either have \lstinline|x = a| and \lstinline|b = y| or \lstinline|x = b| and \lstinline|a = y| modulo \lstinline|us|.
    Assuming that the former holds --- the other case is similar --- we recursively construct the certificates for \lstinline|x = a| and \lstinline|b = y|.
    Together with the fact that \lstinline|a = b|, which we prove by assumption, we obtain \lstinline|x = y| by transitivity.
\end{itemize}

We condense the above intuition into the completeness theorem below.
\begin{lstlisting}
theorem explain_complete:
assumes (x, y) $\in$ equivcl (set us)
shows us $\vdash_=$ explain uf_init us x y : (x, y)
\end{lstlisting}

The \lstinline|explain| function is not sound, though.
This is because it always returns a certificate, even if \lstinline|x| and \lstinline|y| are not equal modulo \lstinline|us|.
To account for this case, we wrap \lstinline|explain| into a partial function \lstinline|explain_partial| (cf.\ \autoref{fig:explain}) that fails if \lstinline|x = y| is not provable.
Soundness and completeness can then be lifted from the soundness of $\vdash_=$ and the completeness of \lstinline|explain|, respectively.
At first glance, it might seem vacuous to define \lstinline|explain_partial| using \lstinline|equivcl|;
however, \lstinline|equivcl| can actually be implemented using union-find operations as the following lemma demonstrated.
\begin{lstlisting}
lemma
defines uf = ufa_unions uf_init us
shows (x, y) $\in$ equivcl (set us) $\leftrightarrow$ x = y $\lor$
  x $\in$ Field (ufa_$\alpha$ uf) $\land$ y $\in$ Field (ufa_$\alpha$ uf) $\land$
  ufa_rep_of uf x = ufa_rep_of uf y
\end{lstlisting}
\todo[inline]{Improve explanation above? Segue?}
